---
title: "Detecting trends in size-at-age with spatio-temporal models."

author: |
  | Christine C. Stawitz^[Corresponding author: christine.stawitz@noaa.gov] $^1$, 
  James T. Thorson $^2$, Paul D. Spencer $^2$, Timothy E. Essington $^3$, Anne D. Hollowed $^2$
  | $^1$ECS Federal in support of NOAA Fisheries,
  $^2$NOAA Alaska Fisheries Science Center,
  $^3$University of Washington School of Aquatic and Fishery Sciences

header-includes:
    - \usepackage{setspace}\doublespacing
    - \usepackage[nomarkers]{endfloat}    
    - \usepackage{lineno}
    - \linenumbers
    
documentclass: book
date: "July 3, 2019"
site: "bookdown::bookdown_site"
output:
  bookdown::pdf_book:
    keep_tex: yes
    toc: no
fig_caption: yes
    

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)
options(xtable.comment = FALSE)
library(bookdown)
require(xtable)
require(here)
require(ggplot2)
require(colorspace)
require(MARSS)
require(broom)
require(plyr)
require(maps)
require(xtable)
source("./R/color_palette.R")

tot_years <- 1982:2018
spp.names <- c("P. cod", "pollock",
               "yellowfin", "N. rock sole", "arrowtooth", "flathead")
bold.somerows <- 
        function(x){ gsub('BEST(.*)',paste('\\\\textbf{\\1','}'),x)}
```

## Abstract {-}
To prepare for climate change, we seek a better understanding of mechanistic links between fish life history processes and climate drivers. If multiple stocks within a large marine ecosystem (LME) exhibit similar temporal patterns, this could indicate an ecosystem-wide response. A shared response to physical processes across species is stronger evidence of a causal link than correlations between temporal variability observed in an individual species' data and physical indices such as sea surface temperature. However, fish populations are often examined at the large marine ecosystem (LME) scale, and spatial variability within LMEs can obfuscate temporal signals of climate response in fisheries data. Here, we estimate trends in body size of six groundfish species within the Bering Sea ecosystem in a model accounting for spatial and spatio-temporal random effects. From the best model chosen for each species (4/6 including spatio-temporal random effects), we extracted and compared estimated interannual size trends between species to identify common patterns. We observed significant positive correlations between the estimated temporal patterns of size for three species (Pacific cod, walleye pollock, and yellowfin sole), and a significant negative correlation between these species and a fourth (flathead sole), so we then used spatial dynamic factor analysis to identify synchrony between extracted trends. Model selection using AICc suggested one shared temporal trend between three species (Pacific cod, walleye pollock, and yellowfin sole) and a negative influence of this trend on a fourth species (flathead sole) best explained these shared patterns. The shared trend towards smaller body size-at-age from 1982-2004 and larger size-at-age from 2005 - 2017 (for Pacific cod, walleye, yellowfin sole) and the opposite trend for flathead sole, could indicate a shared response to ecosystem shifts. However, the trend we observed was not significantly related to mean bottom temperature or the size of the cold pool area of the Bering Sea. Moreover, spatial covariance in size-at-age was species-specific, suggesting local density effects on size-at-age may exist for these species in addition to ecosystem-wide temporal patterns that are shared across species. Accounting for spatial and spatio-temporal variance in fisheries data is important when searching for synchrony in multi-species response to climate. 

## Methods {-}

We analyzed mean length-at-age for a number of Eastern Bering Sea species: Pacific cod (*Gadus macrocephalus*), walleye pollock (*Gadus chalcogrammus*), yellowfin sole (*Limanda aspera*), Northern rock sole(*Lepidopsetta polyxystra*), arrowtooth flounder (*Atheresthes stomas*), and flathead sole(*Hippoglossoides elassodon*). Species were chosen as they are distributed throughout the Eastern Bering Sea large marine ecosystem, have databases of fishery-independent length and age samples dating back to 1982, and are commercially and/or ecologically important to the region. We first fit a single-species spatio-temporal model to average length by year for each species using the VAST (Vector Autoregressive Spatio-Temporal) R package, which allows for simultaneous estimation of spatial and spatio-temporal random effects. We then used spatial Dynamic Factor Analysis to estimate within a multi-species model if there were shared trends between temporal, spatial, or spatio-temporal random effects for each species were accounted for.  

### Data {-}

Fishery-independent length and age data were extracted from the Bering Sea and Aleutian Islands' surveys conducted annually by the National Oceanic and Atmospheric Administration's Alaska Fisheries Science Center. The data extracted included measured length, age, sex, and weight of each sampled fish. Sampling covariates were also extracted for each data point. These covariates include the start latitude, longitude, bottom temperature and depth of each survey trawl tow, in addition to data on the gear type and net width used to collect fish.

### Model structure {-}

We calculated average length-at-age for each year at each unique combination of start latitude and longitude of survey tow. In some years, ages were collected based on length-stratified sampling, which aged equal numbers of fish from each pre-specified length group; this can bias estimates of mean  length-at-age. Length groups were pre-specified based on 1 cm bins (confirm in sampling protocol documents). To obtain accurate estimates of population length-at-age, we corrected for the effect of length-stratified ageing as in Bettoli and Miranda (1992). For each species, we selected the age of fish for which the largest sample size of age data were present. The corrected mean length for each location and each year of data for each species *i* was our input data ($b_{l,s,t}$) where $l$ represents each unique sampling latitude and longitude combination, $s$ represents each species, and $t$ is the year of sampling. 

Each species' data was fit independently using the spatiotemporal modeling framework developed by Thorson and Barnett (2017). First, 100 knots were spatially distributed based on the results of a k-means clustering algorithm to minimize the distance between available data ($b_{l,s,t}$) and the nearest knot. This allowed aggregation of the length observations into $i=100$ values ($b_{i,s,t}$) from the total set of >1600 $l$ sampling stations.

The structure of the models was kept as consistent as possible between species. Data were fit using a Poisson-link model (Thorson 2017) with encounter probabilities fixed at 1 (i.e. because all data were non-zero):
All species' mean length data were fit using a bias-corrected lognormal distribution: 

\begin{equation} 
b_{i,s,t} \sim \mathrm{Lognormal}\left(r-\frac{sigma_m^2}{2}, \sigma_m^2\right)  
\end{equation}


where the mean is calculated as a Poisson link function of the linear predictor $p_{i,s,t}$:
\begin{equation}
r=a_i\mathrm{exp}(p_{i,s,t})
\end{equation}

Where $a_i$ is the area of knot $i$, and $p_{i,s,t}$ is the linear predictor estimated as the following function of annual fixed effects, spatio- and spatio-temporal random effects, and covariates: 
\begin{equation}
p_{i,s,t} = \beta_{s,t}+\sum L_{\omega,s}\omega_{i,s} +I_s\sum L_{\varepsilon,s}\varepsilon_{i,s,t} + \sum_{p=1}^{n}\gamma_{p,i,s,t}\mathbf{X_{p,i,s,t}}
\end{equation}

Where $p_{i,s,t}$ is the mean length for species $s$ in spatial knot $i$ for year $t$, $\beta_{s,t}$ is the fixed effect denoting the year-specific mean length intercept for each species $s$, $L_{\omega}$ is the scalar value representing the absolute standard deviation of the spatial random effect for each species $s$. $I_s$ is an indicator variable denoting whether spatio-temporal variation is included in the model for species $s$, and $L_{\varepsilon}$ is a scalar value reprenting the standard deviation of the spatio-temporal random effect for each species $s$. $\gamma_p$ is the coefficient on the effect of each covariate matrix $X_p$ on mean length, where $n$ is the total number of covariates.

Spatial and spatio-temporal random effects (when included) were modeled using multivariate normal distributions with mean 0 and constant standard deviations for each species and type of random effect (i.e. spatio- or spatio-temporal):
\begin{align}
\omega_s &\sim MVN(0,\mathbf{R_s})\\
\varepsilon_{s,t} &\sim MVN(0,\mathbf{R_s})
\end{align}
Where $R_s$ is the spatial correlation matrix for each species $s$ defined by a Mat\'{e}rn correlation function with smoothness $\phi=1$:

\begin{equation}
R_s(s_n,s_m)=\frac{1}{2^{\phi-1}\Gamma(\phi)}\times(\kappa_s|\mathbf{d_s}(s_n,s_m)\mathbf{H_s})^{\phi}\times K_{\nu}(\kappa_s|\mathbf{d_s}(s_n,s_m)\mathbf{H_s}|)
\end{equation}

where $\kappa_s$ is the decorrelation rate (i.e. how quickly spatial correlation between locations decays with distance), $\mathbf{d_s}(s_n,s_m)$ is the matrix containing distances between locations $s_n$ and $s_m$, $\mathbf{H_s}$ is the anisotropy matrix, and $K_\nu$ is the modified Bessel function of the second kind. 

The anisotropy matrix $\mathbf{H_s}$ specifies the degree to which correlations decline at different rates in different directions.  The Eastern Bering Sea is a continental shelf system with persistent fronts dividing the inner (depth<50m), middle (100m>depth>50m), and outer(depth>100m) domains (Coachman et al. 1986). Thus, we expect correlations to decline more quickly traversing from north to south than along the shelf and accounting for geometric anistropy is likely important in this system. $\mathbf{H_s}$ is estimated as 
\begin{equation}
H=\begin{bmatrix} exp(h_{s,1}) & h_{s,2}  \\ h_{s,2} & exp(-h_{s,1})(1+h_{s,2}^2) \end{bmatrix} \end{equation}

where $h_{s,1}$ and $h_{s,2}$ are the parameters governing geometric anisotropy.

For all species, we first evaluated whether the inclusion of spatio-temporal random effects was supported. We judged this based on two criteria: the magnitude of the estimated spatio-temporal standard deviation (as measured by the loading value $L_{\varepsilon}$) and the AIC values. In all cases where model selection using AIC did not support the inclusion of spatio-temporal effects, the magnitude of the loading value was very small ($L_{\varepsilon}<1.0E-9$). For all species, the magnitude of the estimated spatial standard deviation ($L_{\omega}$) exceeded 0.01, justifying including spatial random effects. We then examined if the data supported the inclusion of covariates by comparing AIC values between models that did and did not include covariates. For all species, we examined the covariates of mean bottom temperature measured during the tow and mean depth of the tow. 

### Model Fitting and Selection {-}
All spatiotemporal models were fit using the VAST R package version 8.0.0, which is described in Thorson and Barnett (2017) and available at http://www.github.com/james-thorson/VAST. All models were fit using a spatial mesh with 50km grid resolution and aggregated into 100 knots across the data area. The model was run for a maximum of 1000 iterations, after which we verified convergence by ensuring no estimated parameter gradient exceeded 1E-3. We also verified observed data conformed to a lognormal distribution by comparing the empirical distribution to a simulated lognormal distribution in a Q-Q plot and a density histogram. We chose the "best spatio-temporal model" for each species as the model with the lowest AIC values, or the simpler of models with $\Delta\mathrm{AIC}<2$ if AIC values were very close. All code used to run the analyses in this paper can be found at http://www.github.com/cstawitz/spatiotemporalBSAIgrowth .

### Dynamic Factor Analysis {-}
We then re-analyzed the data for the four species appearing to have positive or negative temporal correlations in the single-species analysis using a multi-species model. We started with saturated (i.e. four factors for each component) structure on temporal, spatial, and spatio-temporal random effects, then simplified the model until we were able to achieve convergence (measured as the maximum model gradient for an estimated parameter < 1E-04).

Some modifications to the model structure were required to estimate shared factors. First, we switched from using a Poisson-link function to the conventional delta model used in the VAST package. As in the single-species examples described above, we fix the values of the first predictor at 1. This resulted in a change of the predicted mean length at age from the Poisson link in equation (2) above to the equation below:

\begin{equation}
r = a_i \mathrm{log}^{-1}(p_{i,s,t})
\end{equation}

We also used a random walk structure to represent temporal effects in order to support estimating shared temporal factors between species. This had the effect of estimating the $\beta_{s,t}$ parameters described in equation (3) as a random walk instead of independent fixed effects. 
 
## Results {-}

### Single-species spatio-temporal models {-}

The data supported including spatio-temporal random effects for 4 out of 6 stocks, and sampling depth was the external covariate most often found to be influential (3/6 stocks) (Table 1). The model with the lowest AIC for each species included spatio-temporal variation for Pacific cod, walleye pollock, Northern rock sole, and flathead sole. Because all of the estimated spatial variances exceeded 0.01, we did not examine models excluding spatial covariation. The inclusion of environmental covariates was not supported by the data for two of the most data-rich species (Pacific cod, walleye pollock). For the other two species that exhibited spatio-temporal variation, an environmental covariate (sampling depth for Northern rock sole, sampling temperature for flathead sole) was chosen by model selection to be influential. For two species (arrowtooth flounder, yellowfin sole), the variance of spatio-temporal variation was estimated to be smaller than $1E-06$ and was not included; however, model selection supported inclusion of sampling depth as a covariate in the best models for these species. The magnitude of the coefficient on environmental covariates (i.e. depth and temperature) was always <0.2, suggesting these do not have a strong effect on observed length even when their inclusion is supported by model selection. All of the coefficients on covariates included in the best chosen model for each species were positive, e.g. body size is estimated to increase with greater bottom temperature and greater depth.

Predicted spatial patterns of length differed between species. Three species (Pacific cod, Northern rock sole, and yellowfin sole) were predicted to increase in size towards the northwest, following the Bering Sea shelf edge. Walleye pollock showed an opposite pattern, with individuals predicted to be larger-at-age more inshore. The best chosen model for flathead sole predicted larger individuals in the south east, with size decreasing to the north and west. Arrowtooth flounder showed nearly the opposite pattern, with the largest predicted size in the northwest, although temporal variability in size of arrowtooth flounder was much greater than estimated spatial variability. 

Two of the model set chosen to test had poor convergence. A model including spatial covariation and a temperature covariate for arrowtooth flounder did not converge. The model including spatio-temporal covariation and a depth covariate for Northern rock sole produced a positive-definite Hessian, but the maximum gradient had a magnitude of 0.15, suggesting the model may not be converged. Attempting to scale the temperature and depth covariates in alternative ways (i.e. z-scoring, centering) for these two models did not improve convergence.

### Potential synchrony and multispecies DFA {-}

The z-scored prediction of mean length by year from the best model for each species is shown in Figure 1. In these predictions, a number of species (walleye pollock, yellowfin sole, Pacific cod) are shown to have predicted smaller lengths (scaled length < 0) in the period from 1985-1990, while being comparably larger from 2010-2013. A Pearson's correlation test suggested significant correlations between walleye pollock and Pacific cod ($\rho=0.38$), and Pacific cod and yellowfin sole ($\rho=0.39$) (Table 2). For this reason, we used dynamic factor analysis to examine if there was a shared trend between species and if this trend was related to climate drivers. The fully-saturated (four factors each for temporal, spatial, and spatio-temporal covariance) model did not converge (maximum gradient = 3.5E-04). In this model, 0% of the temporal variance was explained by the 3rd and 4th factor, and 0% of the spatial variance was explained by the 4th factor, so we reduced the temporal factor dimensionality to 2 and the spatial factor dimensionality to 3. This model converged (maximum gradient = 9.2E-05). 

The first temporal factor (Figure 2) explained 85.2% of the variance, with positive loadings for walleye pollock (0.0017), Pacific cod (0.026), and yellowfin sole (0.0066) and negative loadings for flathead sole (-0.0081). This trend was negative from 1987 - 2005, with lows  in the early 1990's and 1999-2001, and positive from 2006 - 2017, peaking in the last two years. The second temporal factor explained 14.8% of the variance and had positive loadings for walleye pollock (0.0097) and yellowfin sole (0.0048) and negative loadings for Pacific cod (-0.0030) and flathead sole (-0.0037) (Figure 2). This trend was positive from 1990-onward, with highs from 2006 - 2013.

The first spatial factor explained 70.1% of the variance with positive loadings for yellowfin sole (0.054) and flathead sole (0.12) and negative loadings for Pacific cod (-0.014) and walleye pollock (-0.014) (Figure 3). This spatial pattern was characterized by larger individuals closer to the inshore Aleutian island chain and Bristol bay and a decrease in length-at-age moving northward. The second spatial factor explained 19.8% of the variance, with negative loadings observed for walleye pollock (-0.038) and flathead sole (-0.025) and positive loadings for Pacific cod (0.018) and yellowfin sole (0.050) (Figure 4). This factor was characterized by smaller individuals inshore and larger sizes observed northward and deeper along the Bering Sea shelf edge. The third spatial factor explained 10.2% of the variance and had positive loadings across all species (pollock: 0.025, Pacific cod: 0.043, yellowfin sole: 0.0059, flathead sole: 0.0054). This factor was associated with nearly uniform size-at-age throughout the Bering Sea shelf excluding the typical cold pool area in the North middle domain, which had smaller sizes-at-age (Figure 4).

The first spatiotemporal factor explained 60.2% of the variance, with positive loadings on this factor observed for flathead sole and yellowfin sole (0.0093; 0.080) and negative loadings for Pacific cod and walleye pollock (-0.012; -0.0067; Figure 5). The second spatiotemporal factor explained 19.8% of the variance, with positive loadings for walleye pollock and yellowfin sole (0.016; 0.044) and small negative loadings for Pacific cod and flathead sole (-0.00052, -0.0026; Figure 5). The third spatiotemporal factor explained 16.2% of the variance, with positive loadings observed for walleye pollock, Pacific cod, and flathead sole (0.023; 0.034; 0.0073) and negative loadings for yellowfin sole (-0.0073; Figure 5). The fourth spatiotemporal factor explained 3.7% of the variance, with negative loadings for walleye pollock and flathead sole (-0.015; -0.0020) and positive loadings for Pacific cod and yellowfin sole (0.012; 0.0057). 

```{r echo=FALSE, results="asis"}

folders <- list.dirs(".", recursive=FALSE)
out_folders <- folders[grep("poisson",folders)]
outfiles <- paste0(out_folders,"/Save.RData")
source("./R/create_est_table.R")
out_df<-get_aic_tables(outfiles)
out_df <- out_df[order(out_df$AIC),]
rownames(out_df) <- c("BEST N. rock sole, ST, D", 
                 "N. rock sole, ST",
                 "N. rock sole, S",
                 "N. rock sole, ST, T",
                 "BEST Flathead, ST, T",
                 "Flathead, ST, D",
                 "Flathead, ST",
                 "Flathead, S",
                 "BEST Arrowtooth, S, D",
                 "Arrowtooth, S",
                 "Arrowtooth, ST",
                 "BEST Yellowfin, S, D",
                 "Yellowfin, S",
                 "Yellowfin, S, T",
                 "Yellowfin, ST",
                 "BEST Pollock, ST",
                 "Pollock, ST, D",
                 "Pollock, ST,T",
                  "Pollock, S",
                 "BEST P. cod, ST",
                 "P. cod, ST, T",
                 "P. cod, ST, D",
                 "P. cod, S")
mod.table<- xtable(out_df, digits=3, format="latex", caption="Estimated parameter values, AIC values, and maximum gradient of estimated parameters. Values of 0 in the gamma column denote covariates are not included in this model. Name denotes the model configuration, where 'S' refers to models including spatial covariance only, 'ST' refers to models including spatio-temporal covariance, and 'D' and 'T' denote including tow depth and tow temperature covariates, respectively.")
display(mod.table)[c(1:6,8)] <- "e"
print(mod.table, sanitize.rownames.function=bold.somerows)
```
```{r echo=FALSE, results="asis"}
load("Correlation.RData")
cor_mat <- data.frame(round(cor_mat,3))
cor_mat[] <- lapply(cor_mat, as.character)
cor_mat[1,2] <- paste0("BEST", cor_mat[1,2])
cor_mat[2,3] <- paste0("BEST", cor_mat[2,3])
colnames(cor_mat) = spp.names
rownames(cor_mat) = spp.names[-6]

cor.x<-xtable(cor_mat, format="latex",caption ="Pearson's correlation between scaled length trends. Bold values denote significance at p<0.05 level.")
print(cor.x, sanitize.text.function=bold.somerows)
```


```{r echo=FALSE, results="asis"}
load("DFAAICc.RData")
AICcdf <- data.frame(c("BEST One trend, equal var",
                  "Two trends, equal var",
                  "Three trends, equal var",
                  "One trend, equal var, temp covariate",
                  "One trend, unequal var",
                  "One trend, equal var, cold pool covariate",
                  "Four trends, equal var"),
                sort(AICCvals))

colnames(AICcdf) <- c("Model structure", "AICc")

xaic<-xtable(AICcdf, digits=3, format="latex", caption="AICc values from alternative dynamic factor analysis models fit to scaled length trends.")
display(xaic)[2] <- "e"
print(xaic, include.rownames=FALSE, sanitize.text.function=bold.somerows) 
```


```{r plottrends, echo=FALSE, fig.cap="Scaled weighted annual mean length predictions from best VAST model for Pacific cod (mauve), walleye pollock (brown), yellowfin sole (green), Northern rock sole (teal), arrowtooth flounder (blue), and flathead sole (purple)."}
bestmods <- c("pcod_spatiotemp_poisson", "Pollock_Spatiotemp_Poisson", "yellowfin_spatio_poisson_depth", "nrock_spatiotemp_poisson_depth", "arrow_spatio_poisson_depth", "flathead_spatiotemp_poisson_temp")
source("trendplot.R")
get_trend_plot(bestmods, 1982:2018, spp.names = c("P. cod", "pollock",
                 "yellowfin", "N. rock sole", "arrowtooth", "flathead"))
```

```{r DFAloadings, echo=FALSE, fig.cap="Estimated temporal factors that best describes extracted annual growth trend and loadings of each species' length-at-age trend on each temporal factor. The percentage shown is the percentage of temporal variance explained by each factor across all species."}
source("./R/plot_loadings.R")
old_wd <- getwd()
setwd("./twot_3s_4st_cpa")
best_out<- get(load("Save.RData"))
mapdetails <- get(load("MapDetails.RData"))
factors <- get(load("BestModFactors.RData"))
L_pj <- factors$Rotated_loadings$Beta2
  par(mai = c(1, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0), mfcol=c(2,2))
plot(factors$Rotated_factors$Beta2[,1,]~tot_years, type="l", ylab = "Factor", xlab = "Year", las=1)
abline(h=0)

plot(factors$Rotated_factors$Beta2[,2,]~tot_years, type="l",  ylab = "Factor", xlab = "Year", las=1)
abline(h=0)
plot_loadings(L_pj, tot_years,2)

```
```{r Spatialloadings, echo=FALSE, fig.cap="Estimated spatial factors that best describes extracted annual growth trend and loadings of each species' length-at-age trend on each temporal factor. The percentage shown is the percentage of temporal variance explained by each factor across all species."}
L_pj <- factors$Rotated_loadings$Omega2
par(mai = c(1, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0), mfcol=c(2,2))
#plot(factors$Rotated_factors$Beta2[,1,]~tot_years, type="l", ylab = "Factor", xlab = "Year", las=1)
#abline(h=0)

#plot(factors$Rotated_factors$Beta2[,2,]~tot_years, type="l",  ylab = "Factor", xlab = "Year", las=1)
#abline(h=0)
plot_loadings(L_pj, tot_years,3)
```


\begin{figure}
  \includegraphics[width=\linewidth]{Figures/Factor_maps--Omega2.png}
  \caption{Estimated spatial factors.}
  \label{fig:matdata}
\end{figure}


```{r SpatialTemporalloadings, echo=FALSE, fig.cap="Estimated loadings of each species' length-at-age trend on each spatio-temporal factor. The percentage shown is the percentage of spatio-temporal variance explained by each factor across all species."}

L_pj <- factors$Rotated_loadings$Epsilon2
par(mai = c(1, 0.5, 0.5, 0.1), omi = c(0, 0, 0, 0), mfcol=c(2,2))
#plot(factors$Rotated_factors$Beta2[,1,]~tot_years, type="l", ylab = "Factor", xlab = "Year", las=1)
#abline(h=0)

#plot(factors$Rotated_factors$Beta2[,2,]~tot_years, type="l",  ylab = "Factor", xlab = "Year", las=1)
#abline(h=0)
plot_loadings(L_pj, tot_years,4)
```

\begin{figure}
  \includegraphics[width=\linewidth]{Figures/Factor_maps--Epsilon2.png}
  \caption{Estimated spatiotemporal factors, averaged across years.}
  \label{fig:matdata}
\end{figure}
